---
title: "StarMethod_GTEx"
author: "Songjoon Baek"
format: docx
editor: visual
output: false
warning: false
---

Example

<https://star-protocols.cell.com/protocols/1425>

#### Before you begin

TIMING: 0.5-1h

1.  The hardware specifications for the computing platform used to estimate the timing for each step are provided in the **key resources table**.

2.  The R software environment for statistical computing and graphics is required for this protocol. The latest R version (4.4.1), downloaded from [https://cran.r-project.org](https://cran.r-project.org/), was used to perform the protocol below. This protocol describes the specific steps for analysis of GTEx gene expression data.

3.  The R Studio integrated development environment (IDE) can be downloaded from <https://www.rstudio.com/products/rstudio/>.

4.  R packages used in this protocol are listed under the Software and algorithms heading of the **key resources table.** To install the listed R packages, first install BiocManager with commands:

    `{r}#} chooseCRANmirror() install.packages("BiocManager")`

5.  Then, run the following command to install the R packages listed under the Software and algorithms heading of the **key resources table.**

    `{r}#} BiocManager::install("name_of_package")`

6.  R is always pointed at a designated directory. Specify a working directory at the start of each R session with the following command:

    `{r}#} setwd("path_to_the_desired_folder")`

#### Key resources table

| REAGENT OR RESOURCE | SOURCE | IDENTIFIER |
|------------------------|------------------------|------------------------|
| **Deposited data** |  |  |
| A data dictionary that describes each column in the GTEx_Analysis_v8_Annotations_SampleAttributesDS.txt | [The GTEx Portal](https://www.gtexportal.org/) | [GTEx_Analysis_v8_Annotations_SampleAttributesDD.xlsx](https://storage.googleapis.com/adult-gtex/annotations/v8/metadata-files/GTEx_Analysis_v8_Annotations_SampleAttributesDD.xlsx) |
| A data dictionary that describes each column in the GTEx_Analysis_v8_Annotations_SubjectPhenotypesDS.txt | [The GTEx Portal](https://www.gtexportal.org/) | [GTEx_Analysis_v8_Annotations_SubjectPhenotypesDD.xlsx](https://storage.googleapis.com/adult-gtex/annotations/v8/metadata-files/GTEx_Analysis_v8_Annotations_SubjectPhenotypesDD.xlsx) |
| A de-identified, open access version of the sample annotations available in dbGaP | [The GTEx Portal](https://www.gtexportal.org/) | [GTEx_Analysis_v8_Annotations_SampleAttributesDS.txt](https://storage.googleapis.com/adult-gtex/annotations/v8/metadata-files/GTEx_Analysis_v8_Annotations_SampleAttributesDS.txt) |
| A de-identified, open access version of the subject phenotypes available in dbGaP | [The GTEx Portal](https://www.gtexportal.org/) | [GTEx_Analysis_v8_Annotations_SubjectPhenotypesDS.txt](https://storage.googleapis.com/adult-gtex/annotations/v8/metadata-files/GTEx_Analysis_v8_Annotations_SubjectPhenotypesDS.txt) |
| Gene Expression Transcripts Per Million (TPM) from RNASeQCv2.4.2. | [The GTEx Portal](https://www.gtexportal.org/) | [GTEx_Analysis_2017-06-05_v8_RNASeQCv1.1.9_gene_tpm.gct.gz](https://storage.googleapis.com/adult-gtex/bulk-gex/v8/rna-seq/GTEx_Analysis_2017-06-05_v8_RNASeQCv1.1.9_gene_tpm.gct.gz) |
| **Software and algorithms** |  |  |
| MacOS Sequoia 15.3.2 | Apple Inc. | <https://www.apple.com> |
| R (4.4.1) | The R Project | <https://CRAN.R-project.org/> |
| RStudio (2024.09.1+394) | RStudio Team | <https://www.rstudio.com/products/rstudio/> |
| BiocManager (1.30.25) | [Morgan (2021)](https://star-protocols.cell.com/protocols/1425#bib28) | [https://CRAN.R-project.org/package=BiocManager](https://cran.r-project.org/package=BiocManager) |
| ComplexHeatmap |  |  |
| colorRamp2 |  |  |

## Materials and equipment

The analyses outlined in this protocol utilize publicly available RNA-Seq clinical datasets and several well-established CRAN and BioConductor R packages. All human subjects have been de-identified. The ethical principles set forth by the International Ethical Guidelines for Health-related Research Involving Humans ([CIOMS, 2016](https://star-protocols.cell.com/protocols/1425#bib9)) should preside over any activities that fall under health-related research with humans, such as biobanking and epidemiological studies. Researchers who study these data should also comply with TCGA and GTEx policies such as maintaining participants’ privacy, accessing data securely, and following TCGA and GTEx publication guidelines. To cite R packages, run the following command in R.

## Step-by-step method details

#### Data import, cleaning, and preprocessing

TIMING: 1-4h

1.  Download the data dictionaries and de-identified sample annotations files (see the Key Resources Table for the link), and save them in the "data" folder within your working directory.
2.  Next, load these files into the R environment. For convenience, renamed the row names of the **pheno.ds** object to use subject IDs.

```{r}
#| message: false
#| warning: false
attr.dd = readxl::read_xlsx("GTEx_Analysis_v8_Annotations_SampleAttributesDD.xlsx");
pheno.dd = readxl::read_xlsx("GTEx_Analysis_v8_Annotations_SubjectPhenotypesDD.xlsx");
attr.ds = read.delim("GTEx_Analysis_v8_Annotations_SampleAttributesDS.txt");
pheno.ds = read.delim("GTEx_Analysis_v8_Annotations_SubjectPhenotypesDS.txt");
row.names(pheno.ds) = pheno.ds$SUBJID;
```

3.  Select RNA-seq experiments from the attribution table and create a combined table of RNA-seq experiements.

```{r}
rna_seqs= attr.ds[attr.ds$SMAFRZE == "RNASEQ",1:17];  #17382 out of 22951
getpid = function(x) { substring(x,1,10);}
pids = sapply(rna_seqs$SAMPID, getpid);
combinedtab=cbind(do.call("rbind", lapply(pids, function(x) { pheno.ds[x,];})), rna_seqs);
combinedtab=combinedtab[!is.na(combinedtab$AGE),];   # 12289
head(combinedtab[,c("SAMPID","AGE","SEX")]);
```

```{r}
saveRDS(combinedtab, file="GTEx_Analysis_RNASEQ_samples.Rda")
```

To reduce data-loading time, read the TPM table containing gene expression data and save it in R's internal data format.

```{r}
if (!file.exists("tpm_gct.dat")) {
  tpm.gct <- read.delim(file="GTEx_Analysis_2017-06-05_v8_RNASeQCv1.1.9_gene_tpm.gct.gz", skip=2);
  save(tpm.gct, file="tpm_gct.dat");
} else {
  load("tpm_gct.dat");  
}
```

Afterward, read the saved RNA-seq samples and add a group category based on the age:

- Young: 20-39 years

- Middle-aged: 40-59 years

- Old: 60 years or older

```{r}
#xlsx::write.xlsx(combinedtab, "GTEx_Analysis_RNASEQ_samples.xlsx");  # takes too long
dattab=readRDS(file="GTEx_Analysis_RNASEQ_samples.Rda")
dattab$AGE = factor(dattab$AGE, sort(unique(dattab$AGE)));  # 20-29, 30-39,40-49,50-59,60-69,70-79
age_group = list("20-29"="young","30-39"="young", "40-49"="middle", "50-59"="middle", "60-69"="old", "70-79"="old");
tissue = "Lung";

tab = subset(dattab, dattab$SMTS==tissue)
tab = cbind(tab, group=sapply(tab$AGE, function(x) { age_group[[x]];} ));
tab$group = factor(tab$group, c("young","middle","old"));
```

Use the **summary()** command on the **AGE** variable to examine the age distribution.

```{r}
summary(dattab$AGE)
#20-29 30-39 40-49 50-59 60-69 70-79 
#  804   958  1520  3782  4651   574 
```

#### Gene Selection and Normalization

Timing:1-2 hours

Create a list of genes of interest along with their corresponding categories. The *category* should be a categorical variable indicating the group or functional classification of each gene.

```{r}
IKP = c('CENPN','CENPL','CENPO','CENPQ','CENPU','CENPH','CENPI','CENPK',
        'CENPM','CENPS','CENPT','CENPW','CENPX');
CentP = c('CENPA','CENPC','CENPB');
OKP = c('MIS12','DSN1','NSL1','PMF1','NDC80','NUF2','KNL1','SPC24','SPC25', 'ZWINT');
other=c('CENPE','CENPF','CENPJ','CENPV');
chaperones=c('DAXX','HIRA','HJURP');
selected_genes = c(IKP, CentP, OKP, other, chaperones);
category=factor(c(rep("InnerKinetochore", length(IKP)), 
            rep("Centromere", length(CentP)), 
            rep("OuterKinetochore", length(OKP)), 
            rep("Other", length(other)), 
            rep("Chaperone", length(chaperones))),
            levels = c("InnerKinetochore",
                       "Centromere",
                       "OuterKinetochore",
                       "Other",
                       "Chaperone"));
names(category) = selected_genes;
```

In this analysis, we include a few housekeeping genes, which will later be used for normalization.

```{r}
house_keeping_genes = c("ACTN1","GAPDH","ACTB")
genes_to_use=c(house_keeping_genes,selected_genes);
genes_to_use
```

Generate a gene expression table using the selected genes. Before doing so, remove any genes with low expression to improve data quality.

```{r}
sampid = gsub("-",".", tab$SAMPID) # convert - into . to match the column names
gene_exp = t(tpm.gct[match(genes_to_use,tpm.gct$Description), match(sampid, colnames(tpm.gct))]);
colnames(gene_exp) <- genes_to_use;
gene_exp= gene_exp[,apply(gene_exp,2,sum) > 1];   # delete genes with very low expressions
```

Optionally, for bulk RNA-seq TPM matrices, an alternative filtering strategy is to retain genes with TPM ≥1 in at least 10-20% of samples, helping to reduce the influence of low-level, sporadic expression noise in large sample cohorts.

```{r}
min_tpm <- 1
min_fraction <- 0.10
gene_exp = t(tpm.gct[match(genes_to_use,tpm.gct$Description), match(sampid, colnames(tpm.gct))]);
colnames(gene_exp) <- genes_to_use;
gene_exp <- gene_exp[rowMeans(gene_exp >= min_tpm, na.rm = TRUE) >= min_fraction,]
```

Choose and run only one of the three normalization methods belows:

**Option 1** Normalization by One Housekeeping Gene

Convert the sample ID (SAMPID) in the table, identify the matching TPM expression values, and construct a normalized gene expression table.

```{r}
normalizationByOneGene<-function(gene_exp, gene="ACTN1") {
  normalized = gene_exp;  # 385 x 27
  mean_actin=mean(gene_exp[,gene]);
  for (ii in 1:ncol(gene_exp)) {
    normalized[,ii] = gene_exp[,ii] / gene_exp[,gene]*mean_actin;
  }
  normalized;
}

normalized = normalizationByOneGene(gene_exp, gene="ACTN1")
head(normalized[,1:5])
```

**Option 2** Normalization by a Group of Housekeeping Genes

Use the average or median expression of a defined group of housekeeping genes for normalization.

```{r}
normalizationByGenes<-function(gene_exp, genes=c("ACTN1","GAPDH","ACTB")) {
  normalized = gene_exp;  # 385 x 27
  mean_exp=apply(gene_exp[,colnames(gene_exp) %in% genes],2,mean);
  gene_exp_hk=gene_exp[,colnames(gene_exp) %in% genes] 
  for (i in 1:nrow(gene_exp_hk)) { 
    gene_exp_hk[i,] = gene_exp_hk[i,]  / mean_exp 
  } 
  nfactor  = apply(gene_exp_hk,1,mean) # 385
  for (ii in 1:ncol(gene_exp)) {
    normalized[,ii] = gene_exp[,ii] / nfactor[ii]
  }
  normalized;
}
normalized = normalizationByGenes(gene_exp, genes=c("ACTN1","GAPDH","ACTB"))
head(normalized[,1:5])
```

**Option 3** Normalization by the Upper 75% Quantile

Normalize expression values based on the 75th percentile of gene expression levels.

```{r}
normalizationByQ75 <- function(gene_exp) {
  q75 = apply(gene_exp,1, function(x) quantile(x,0.75))
  mean_q75 = mean(q75)
  nfactor = q75 / mean_q75
  normalized = gene_exp;
  for (ii in 1:ncol(gene_exp)) {
    normalized[,ii] = gene_exp[,ii] / nfactor[ii]
  }
  normalized;
}
normalized = normalizationByQ75(gene_exp);
head(normalized[,1:5])
```

Use the function **getNormalizedExpressionForGenes** to combine the normalized gene expression data with each sample's age and age group. Save the resulting table as an RDS file.

```{r}
getNormalizedExpressionForGenes<-function(geneSymbols, normalized, datainfo) {
  gindex=sapply(geneSymbols, function(x) {which( x== fpkm$gene_name);})
  nfpkm = t(normalized[gindex,]);
  colnames(nfpkm) = fpkm$gene_name[gindex];
  age = round(-as.numeric(datainfo$DTB)/365.2422,1);
  group = factor(datainfo$group, c("young","middle","old"));
  gene.exp = data.frame(group=group, age=age, nfpkm);
  gene.exp;
}

cenpexpr = data.frame(id=tab$SAMPID,age=tab$AGE, group=tab$group, normalized);
saveRDS(cenpexpr, file=sprintf("GTEx_%s_CENP_gene_expr_p1.Rda", tissue));
```

Use the command below to display the available tissues in the data set.

```{r}
unique(dattab$SMTS)
```

To summarize the previous steps, the function **make_GTEx_GeneExpTableForTissue** generates a normalized gene expression table for the selected genes in each tissue and saves as an RDS file.

```{r}
make_GTEx_GeneExpTableForTissue<-function(tissue, genes_to_use) {
  dattab=readRDS(file="GTEx_Analysis_RNASEQ_samples.Rda")
  dattab$AGE = factor(dattab$AGE, sort(unique(dattab$AGE)));  # 20-29, 30-39,40-49,50-59,60-69,70-79
  age_group = list("20-29"="young","30-39"="young", "40-49"="middle", "50-59"="middle", "60-69"="old", "70-79"="old");

  tab = subset(dattab, dattab$SMTS==tissue)
  tab = cbind(tab, group=sapply(tab$AGE, function(x) { age_group[[x]];} ));
  tab$group = factor(tab$group, c("young","middle","old"));

  sampid = gsub("-",".", tab$SAMPID) # convert - into . to match the column names
  gene_exp = t(tpm.gct[match(genes_to_use,tpm.gct$Description), match(sampid, colnames(tpm.gct))]);
  colnames(gene_exp) <- genes_to_use;

  normalized = normalizationByOneGene(gene_exp, gene="ACTN1")
  cenpexpr = data.frame(id=tab$SAMPID,age=tab$AGE, group=tab$group, normalized);

  saveRDS(cenpexpr, file=sprintf("GTEx_%s_CENP_gene_expr_p1.Rda", tissue));
  summary(cenpexpr$group);
}

heart=make_GTEx_GeneExpTableForTissue("Heart",genes_to_use);
lung=make_GTEx_GeneExpTableForTissue("Lung",genes_to_use);
liver=make_GTEx_GeneExpTableForTissue("Liver",genes_to_use);
skin=make_GTEx_GeneExpTableForTissue("Skin",genes_to_use);
colon=make_GTEx_GeneExpTableForTissue("Colon",genes_to_use);
kidney=make_GTEx_GeneExpTableForTissue("Kidney",genes_to_use);
```

``` r
``
```

```{r}
counts0 = rbind(heart=heart, lung=lung, liver=liver, skin=skin, colon=colon, kidney=kidney)
counts1  = cbind(counts0, total=apply(counts0,1,sum))
counts = rbind(counts1, total=apply(counts1,2,sum))
counts
```

Next, merge the gene expression data from six different tissues to create a combined expression table.

```{r}
counttable = rbind(heart=heart,lung=lung, liver=liver, skin=skin, colon=colon, kidney=kidney);
dt = data.frame(tissue=row.names(counttable), counttable);
saveRDS(dt, "gene_exp_heatmap_count_table.rds");
```

#### Z-score and heatmaps

TIMING: 1-4h

For each tissue, we calculated the mean gene expression values grouped by age category. These values were then combined into a single table and reorganized by tissue name and age group for downstream analysis.

```{r}
tissues = c("Kidney","Lung","Liver","Colon","Heart","Skin");
cenpaexpfiles = sprintf("GTEx_%s_CENP_gene_expr_p1.Rda", tissues)
tabs <- lapply(cenpaexpfiles, readRDS);
names(tabs) = tissues;

groups = c("young","middle","aged");
ntissues = length(tissues)
meanexp <- lapply(tabs, function(tab) {
  group_mean <- aggregate(x=tab[,c(-1,-2,-3)], by= list(tab$group), FUN=mean);
  row.names(group_mean) = group_mean[,1];
  group_mean[,-1];
});

summary_tab = data.frame(matrix(0, nrow=3*length(tissues), ncol=ncol(tabs[[1]])-3));

for (j in 1:3) {
  for (i in seq_along(tissues)) {
    summary_tab[(j-1)*length(tissues)+i,]=meanexp[[i]][j,]
    row.names(summary_tab)[(j-1)*length(tissues)+i] = sprintf("%s.%s", tissues[i], groups[j]);
  }
}
colnames(summary_tab) = colnames(tabs[[1]])[c(-1,-2,-3)];
meantab = t(summary_tab);

mtab = meantab[which(!row.names(meantab) %in% c("ACTN1","GAPDH","ACTB")),c(1,7,13,2,8,14,3,9,15,4,10,16,5,11,17,6,12,18)];  # rearranged by tissues
colnames(mtab);
```

```{r}
head(mtab[,1:6])
```

From the gene expression tables of each tissue, calculate z-scores separately for each tissue and combined the results into a single table for heatmap visualization.

```{r}
mtabs = list(Kidney=mtab[,1:3], Lung =mtab[,4:6], Liver = mtab[,7:9], Colon = mtab[,10:12], Heart=mtab[,13:15], Skin=mtab[,16:18]);
calcZscore<-function (tab) {
  mm= apply(tab,1,mean); ssd = apply(tab,1,sd);
  zscore = (tab - mm) / ssd;
  zscore;
}
z.score = do.call("cbind",lapply(mtabs, calcZscore))[selected_genes,];
head(z.score[,1:6])
```

The z-score heatmap for an individual tissue can be generated using the following commands.

```{r}
z.score.liver = z.score[,c("Liver.young","Liver.middle","Liver.aged")]

library(ComplexHeatmap);
library(colorRamp2)

col_fun = colorRamp2(c(-1.5, 0, 1.5), c("blue", "white", "red"));

ha = HeatmapAnnotation(
    age = factor(c("young","middle","aged"),levels=c("young","middle","aged")),  
    col = list(age = c("young" = "#440154FF" , "middle" = "#21908CFF", "aged" = "#FDE725FF") )
)

va = rowAnnotation(
    category = category,
    col = list(category = c("Centromere"="#00DDCF","InnerKinetochore"="#000004FF", 
                       "OuterKinetochore"="#BB3754FF","Other"="#F98C0AFF", "Chaperone"="#CCCC84FF"))
); 

Heatmap(z.score.liver, name = "z-score", col = col_fun,  cluster_columns = FALSE, bottom_annotation = ha,    left_annotation = va,    column_title = "Liver", row_title="Genes")
```

Next, assign specific color schemes to represent z-score levels, tissue types and gene categories. The first heatmap displays gene expression changes across age groups, with genes hierarchically clustered based on their expression patterns.

```{r}
col_fun = colorRamp2(c(-1.5, 0, 1.5), c("blue", "white", "red"));
split = data.frame(factor(c(rep(c("Kidney"), 3),rep(c("Lung"), 3),
                            rep(c("Liver"), 3),rep(c("Colon"), 3), rep(c("Heart"), 3),rep(c("Skin"),3)                   ), levels=c("Kidney","Lung","Liver","Colon","Heart","Skin")));
ha = HeatmapAnnotation(tissues = factor(c(rep(c("Kidney"), 3),rep(c("Lung"), 3), rep(c("Liver"), 3), rep(c("Colon"), 3), rep(c("Heart"), 3),rep(c("Skin"), 3)), levels=c("Kidney","Lung","Liver","Colon","Heart","Skin")), age = factor(c(rep(c("young","middle","aged"),6)),levels=c("young","middle","aged")),col = list(tissues = c("Kidney"="#FF0000", "Lung"="#FFFF00",              "Liver"="#00FF00","Colon"="#00FFFF", "Heart"="#0000FF", "Skin"="#FF00FF"),
                                                                                                                                                                                                                                                                                                                           age = c("young" = "#440154FF" , "middle" = "#21908CFF", "aged" = "#FDE725FF") ))
va = rowAnnotation(category = category,col = list(category = c("Centromere"="#00DDCF","InnerKinetochore"="#000004FF","OuterKinetochore"="#BB3754FF","Other"="#F98C0AFF", "Chaperone"="#CCCC84FF"))); 
Heatmap(z.score, name = "z-score", col = col_fun,  column_split= split, column_gap= unit(0.1, "in"),cluster_columns = FALSE, bottom_annotation = ha,
        left_annotation = va,column_title = "Tissues / Age Groups", row_title="Genes")


```

```{r}
library(ComplexHeatmap);
library(colorRamp2)
#library(circlize)
col_fun = colorRamp2(c(-1.5, 0, 1.5), c("blue", "white", "red"));
split = data.frame(factor(c(rep(c("Kidney"), 3),
                   rep(c("Lung"), 3),
                   rep(c("Liver"), 3),
                   rep(c("Colon"), 3),
                   rep(c("Heart"), 3),
                   rep(c("Skin"), 3)
                   ), levels=c("Kidney","Lung","Liver","Colon","Heart","Skin")));

ha = HeatmapAnnotation(
    tissues = factor(c(rep(c("Kidney"), 3),
                   rep(c("Lung"), 3),
                   rep(c("Liver"), 3),
                   rep(c("Colon"), 3),
                   rep(c("Heart"), 3),
                   rep(c("Skin"), 3)
                   ), levels=c("Kidney","Lung","Liver","Colon","Heart","Skin")),
    age = factor(c(rep(c("young","middle","aged"),6)),levels=c("young","middle","aged")),  
    col = list(tissues = c("Kidney"="#FF0000", "Lung"="#FFFF00",
                       "Liver"="#00FF00","Colon"="#00FFFF", "Heart"="#0000FF", "Skin"="#FF00FF"),
               age = c("young" = "#440154FF" , "middle" = "#21908CFF", "aged" = "#FDE725FF") )
)

va = rowAnnotation(
    category = category,
    col = list(category = c("Centromere"="#00DDCF","InnerKinetochore"="#000004FF", 
                       "OuterKinetochore"="#BB3754FF","Other"="#F98C0AFF", "Chaperone"="#CCCC84FF"))
); 


Heatmap(z.score, name = "z-score", col = col_fun,  column_split= split, column_gap= unit(0.1, "in"),
        cluster_columns = FALSE, bottom_annotation = ha,
        left_annotation = va,
    column_title = "Tissues / Age Groups", row_title="Genes")
```

The second heamap groups genes according to their their predefined categories, allowing for category-based comparison.

```{r}
splitrow= data.frame(category);
Heatmap(z.score, name = "z-score", col = col_fun,  column_split= split, column_gap= unit(0.1, "in"),
        row_split=splitrow, 
        row_gap=unit(0.1,"in"),
        cluster_columns = FALSE,
        cluster_rows= FALSE, 
        bottom_annotation = ha,
        left_annotation = va,
    column_title = "Tissues / Age Groups", row_title="Genes")
```

The third heatmap uses hierarchical clustering of z-scores based on tissue type and age group combinations, highlighting overall trends in gene expression changes with aging.

```{r}
Heatmap(z.score, name = "z-score", col = col_fun, 
        cluster_columns = T, bottom_annotation = ha,
            row_split=splitrow,         left_annotation = va,
    column_title = "Tissues / Age Groups", row_title="Genes")
```

The commands below generate the fourth heatmap, with the columns sorted by age groups.

```{r}
z.score2 = z.score[,c(1,4,7,10,13,16,2,5,8,11,14,17,3,6,9,12,15,18)]

ha2 = HeatmapAnnotation(
    tissues = factor(rep(c("Kidney","Lung","Liver","Colon","Heart","Skin"), 3), levels=c("Kidney","Lung","Liver","Colon","Heart","Skin")),
    age = factor(c(rep("young",6),rep("middle",6),rep("aged",6)),levels=c("young","middle","aged")),  
    col = list(tissues = c("Kidney"="#FF0000", "Lung"="#FDE725FF",
                       "Liver"="#00FF00","Colon"="#00FFFF", "Heart"="#0000FF", "Skin"="#FF00FF"),
               age = c("young" = "#440154FF" , "middle" = "#21908CFF", "aged" = "#FDE725FF")))

splitrow= data.frame(category);
split2= data.frame(factor(c(rep(c("young"), 6),
                   rep(c("middle"), 6),
                   rep(c("aged"), 6)                ), levels=c("young","middle","aged")));


Heatmap(z.score2, name = "z-score", col = col_fun,  column_split= split2, column_gap= unit(0.1, "in"),
        row_split=splitrow, 
        row_gap=unit(0.1,"in"),
        cluster_columns = FALSE,
        cluster_rows= FALSE, 
        bottom_annotation = ha2,
        left_annotation = va,
    column_title = "Tissues / Age Groups", row_title="Genes")
```
